Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2025-07-09 manuscript/manuscript.Rmd
edits to the revised manuscript

As a simple illustration, Figure 1 summarises simulated data for one subject that shows how comparing means between two conditions can conceal the shapes of the underlying RT and accuracy distributions. Indeed, compared to the aggregation of data across trials (Figure 1A), a distributional approach offers the possibility to reveal the time course of psychological states (Figure 1B). For example, Figure 1B shows a first state (up to 400 ms after target onset) for which the early upswing in hazard is equal for both conditions, and the emitted responses are always correct in condition 1 and always incorrect in condition 2. In a second state (400 to 500 ms), hazard is higher in condition 1, and conditional accuracies are close to .5 in both conditions. In a third state (>500 ms), the effect disappears in hazard, and all conditional accuracies are equal to 1. Importantly from a face-validity perspective, this pattern of simulated data can be seen in the experimental psychology literature [@panisWhatShapingRT2016].

(ref:descr-fig1-caption) Simulated single-subject data showing mean performance versus distributional (EHA/SAT) analyses. (A) The mean RT (top) and overall accuracy (bottom) for two conditions are plotted. Two hundred trials are simulated in each condition. (B) The discrete-time hazard functions (top) and conditional accuracy functions (bottom) are plotted for the same data. The first second after target stimulus onset (time zero) is divided in ten bins of 100 ms (t = 1 to 10). The first bin is (0,100], the last bin is (900,1000]. Note that the hazard and conditional accuracy estimates are plotted at the endpoint of each time bin. The definitions of discrete-time hazard and conditional accuracy are further explained in section 2.1.2. Error bars represent +/- 1 standard error of the mean (A) or proportion (B).

@@

To apply discrete-time EHA, one divides the within-trial time in discrete, contiguous time bins indexed by t (e.g., t = 1 to 10; Figure 1B).
Then let RT be a discrete random variable denoting the rank of the time bin in which a particular person’s response occurs in a particular trial (i.e., repeated measure). 
For example, a response in one trial might occur at 546 ms and it would be in time bin 6 (any RTs from 501 ms to 600 ms).
One then calculates the sample-based estimate of the discrete-time hazard function of event occurrence for each experimental condition (Figure 1B upper panel).
The discrete-time hazard function gives you, for each time bin, the conditional probability that the event occurs (sometime) in
bin t, given that the event does not occur in previous bins. 
In other words, it reflects the instantaneous risk that the event occurs in the current bin t, given that it has not yet
occurred in the past, i.e., in one of the prior bins (t-1, t-2, ..., 1).

@@

<!-- Our list of tutorials is as follows:

* 1a. Wrangle raw data and calculate descriptive stats for one independent variable
* 1b. Wrangle raw data and calculate descriptive stats for two independent variables
* 2a. Bayesian multilevel modeling for h(t)
* 2b. Bayesian multilevel modeling for ca(t)
* 3a. Frequentist multilevel modeling for h(t)
* 3b. Frequentist multilevel modeling for ca(t)
* 4.  Simulation and power analysis for planning experiments-->

@@

In general, it is important to visually inspect the functions first for each participant, in order to identify individuals that may not be following task instructions (e.g., a flat conditional accuracy function at .5 indicates that someone is just guessing), outlying individuals, and/or different groups with qualitatively different behavior. Also, to select a suited bin width for model fitting, one can test and compare various bin widths in the censor function, and select the smallest one that is supported by the data. Too small bin widths will result in erratic hazard functions because many bins will have estimates equal to zero.

@@

The middle column of Supplementary Figure 3 (section F of the Supplemental Material) shows six examples of prior distributions for an intercept on the logit and/or cloglog scales. While a normal distribution with relatively large variance is often used as a weakly informative prior for continuous dependent variables, rows A and B of Supplementary Figure 3 show that specifying such distributions on the logit and cloglog scales actually leads to rather informative distributions on the original probability scale, as most mass is pushed to probabilities of 0 and 1.

@@

Estimating model M1i took about 124 minutes.

@@

<!-- , and for each participant in the data set (Figure 4B).-->

<!-- (ref:descr-plot4-caption) Point (median) and 80/95% credible interval summaries of the hazard estimates (expected values of the draws from the posterior predictive distributions) in each time bin at the population level (A), and for each participant (B).-->

```{r plot-pred-grand-effects, fig.cap = "(ref:descr-plot4-caption)", out.width='80%'}
#knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_pred_combined.png")
```

<!-- As we are actually interested in the effects of congruent and incongruent primes, relative to the blank prime condition, we can construct two contrasts (congruent-blank, incongruent-blank), and plot the posterior distributions of these contrast effects at the population level (Figure 3C).-->

<!-- and at the participant level (Figure 5B). -->

<!-- (ref:descr-plot5-caption) Point (mean) and 80/95% credible interval summaries of estimated differences in hazard in each time bin at the population level (A), and for each participant (B).-->

```{r plot-grand-ame-effects, fig.cap = "(ref:descr-plot5-caption)", out.width='80%'}
#knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_ame_combined.png")
```

@@

<!--If we want to focus more on inter-individual differences, we can study the subject-specific hazard functions in Figure 5B. Note that three participants (1, 2, and 3) show a negative difference for the contrast "congruent minus incongruent" in bin (360,400] -- subject 2 also in bin (320,360]. Interestingly, all subjects show a tendency in their mean difference (congruent minus blank) to "dip" around that time. Therefore,-->

<!-- Future studies could (a) increase the number of participants to estimate the proportion of "dippers" in the subject population, and/or (b) try to explain why this dip occurs. For example, @panisWhatShapingRT2016 concluded that active, top-down, task-guided response inhibition effects emerge around 360 ms after the onset of the stimulus following the prime (here: the target stimulus). Such a top-down inhibitory effect might exist in our priming data set, because after some time participants will learn that the first stimulus is not the one they have to respond to. To prevent a premature overt response to the prime they thus might gradually increase a global response threshold during the remainder of the experiment, which could result in a lower hazard in congruent trials compared to blank trials, for bins after ~360 ms, and towards the end of the experiment. This effect might be masked for incongruent primes by the response competition effect.-->

@@

<!--, and for each participant in the data set (Figure 7B).-->

<!-- (ref:plot7-caption) Point (median) and 80/95% credible interval summaries of the conditional accuracy estimates (expected values of the draws from the posterior predictive distributions) in each time bin at the population level (A), and for each participant (B).-->

```{r plot-pred-grand-effects-ca, fig.cap = "(ref:plot7-caption)", out.width='80%'}
#knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_ca_pred_combined.png")
```

<!-- and for each participant (Figure 8B).-->

<!-- (ref:descr-plot8-caption) Point (mean) and 80/95% credible interval summaries of estimated differences in conditional accuracy in each time bin at the population level (A), and for each participant (B).-->

```{r plot-ca-grand-ame-effects, fig.cap = "(ref:descr-plot8-caption)", out.width='80%'}
#knitr::include_graphics("../Tutorial_2_Bayesian/figures/M1i_ca_ame_combined.png")
```

<!-- ## 4.5 Tutorial 3a: Fitting Frequentist hazard models -->

<!-- In this fifth tutorial we illustrate how to fit a multilevel regression model to RT data in the frequentist framework, for the data used in Tutorial 1a. The general process is similar to that in Tutorial 2a, except that there are no priors to set.  -->

<!-- Again, to keep the tutorial concise, we only fit model M1i (see Tutorial 2a) using the function glmer() from the R package lme4. Alternatively, one could also use the function glmmPQL() from the R package MASS [@ripleyMASSSupportFunctions2024]. The resulting hazard model is called M1i_f with the appended "_f" denoting a frequentist model. -->

<!-- In Figure 11 we compare the parameter estimates from the Bayesian regression model M1i with those from the frequentist model M1i_f. -->

<!-- (ref:descr-plot11-caption) Parameter estimates for model M1i from brm() -- means and 95% credible intervals -- and model M1i_f from glmer() -- maximum likelihood estimates and 95% confidence intervals. -->

<!-- ```{r plot-comparison, fig.cap = "(ref:descr-plot11-caption)", out.width='80%'} -->
<!-- knitr::include_graphics("../Tutorial_3_Frequentist/comparison.png") -->
<!-- ``` -->

<!-- Figure 11 confirms that the parameter estimates from both Bayesian and frequentist models are pretty similar, which makes sense given the close similarity in model structure. However, model M1i_f did not converge and resulted in a singular fit. This is of course one of the reasons why Bayesian modeling has become so popular in recent years. But the price you pay for being able to fit models with more complex varying effects structures via a Bayesian framework is increased computation time. In other words, as we have noted throughout, some of the Bayesian models in Tutorials 2a took several hours to build. -->

<!-- ## 4.6 Tutorial 3b: Fitting Frequentist conditional accuracy models -->

<!-- In this sixth tutorial we illustrate how to fit a multilevel regression model to the timed accuracy data in the frequentist framework, for the data used in Tutorial 1a. To be concise, we only fit effects from model M1i_ca (see Tutorial 2b) using the function glmer() from the R package lme4. Alternatively, one could also use the function glmmPQL() from the R package MASS [@ripleyMASSSupportFunctions2024]. The resulting conditional accuracy model, which we labelled M1i_ca_f, did not converge and resulted in a singular fit. Again, this just highlights some of the difficulties in fitting reasonably complex varying/random effects structures in frequentist workflows. -->

@@

The general approach to planning that we adopt here involves simulating reasonably structured data to help guide what you might be able to expect from your data once you collect it [@gelmanBayesianWorkflow2020].
The basic structure and code follows the examples outlined by Solomon Kurz in his 'power' blog posts (https://solomonkurz.netlify.app/blog/bayesian-power-analysis-part-i/) and Lisa Debruine's R package faux{} (https://debruine.github.io/faux/) as well as these related papers [@debruineUnderstandingMixedEffectsModels2021; @pargentTutorialTailoredSimulationBased2024].

@@

Summary results from Simulation 1 are shown in Figure 6A.
Figure 6A depicts statistical "power" as calculated by the percentage of lower-bound 95% confidence intervals that exclude zero when the difference between prime condition is calculated (congruent - incongruent). 
In other words, what fraction of the simulated datasets generated an effect of prime that excludes the criterion mark of zero.
We are aware that "power" is not part of a Bayesian analytical workflow, but we choose to include it here, as it is familiar to most researchers in experimental psychology.

@@

**But**, and this is an important "but", unless there are unavoidable reasons, no matter what planning choices we made based on these data simulations, we would not solely rely on data collected from one single study. 
Instead, we would run a follow-up experiment that replicates and extends the initial result. 
By doing so, we would aim to avoid the Cult of the Isolated Single Study [@nelderStatisticsStatisticalScience1999; @tongStatisticalInferenceEnables2019], and thus reduce the reliance on any one type of planning tool, such as a power analysis. 
Then, we would look for common patterns across two or more experiments, rather than trying to make the case that a single study on its own has sufficient evidential value to hit some criterion mark.

@@

For those researchers, like ourselves, who are primarily interested in understanding human cognitive and brain systems, we consider two broadly-defined, main use-cases of EHA. First, as we hope to have made clear by this point, EHA is one way to investigating a "temporal states" approach to cognitive processes, by tracking behavior as a function of step-wise increases in absolute waiting time. EHA thus provides a way to uncover the microgenesis of cognitive effects, by revealing when cognitive states may start and stop, how states are replaced with others, as well as what they may be tied to or interact with. Therefore, if your research questions concern **when psychological states occur, and how they are temporally organized**, our EHA tutorials could be useful tools for you to use.

Second, even if you are not primarily interested in studying the temporal organization of cognitive states, EHA could still be a useful tool to consider using, in order to qualify inferences that are being made based on comparisons between means. Given that distinctly different inferences can be made from the same data based on whether one computes a mean across trials or a RT distribution of events (Figure 1), it may be important for researchers to supplement comparisons between means with EHA.
In any case, if you have a lot of right-censored observations in your RT data set, and/or your research question concerns whether and when responses occur, and whether and when experimental manipulations affect the instantaneous risk of response occurrence, then EHA should be your method of choice.

<!--One could envisage scenarios where the implicit assumption of an effect in mean RTs manifesting across all of the time bins measured would not be supported by EHA. Therefore, the conclusion of interest would not apply to all responses, but instead it would be restricted to certain periods of within-trial time.-->

>>>>

As a simple illustration, Figure 1 summarises simulated data for one subject that shows how comparing means between two conditions can conceal the shapes of the underlying RT and accuracy distributions. Indeed, compared to the aggregation of data across trials (Figure 1A), a distributional approach known as event history analysis (EHA) offers the possibility to reveal the time course of psychological states (Figure 1B). For example, Figure 1B shows a first state (up to 400 ms after target onset) for which the early upswing in hazard is equal for both conditions, and the emitted responses are always correct in condition 1 and always incorrect in condition 2. In a second state (400 to 500 ms), hazard is higher in condition 1, and conditional accuracies are close to .5 in both conditions. In a third state (>500 ms), the effect disappears in hazard, and all conditional accuracies are equal to 1. Importantly from a face-validity perspective, this pattern of simulated data can be seen in the experimental psychology literature [@panisWhatShapingRT2016].

(ref:descr-fig1-caption) Simulated single-subject data showing mean performance versus a distributional analysis. (A) The mean RT (top) and overall accuracy (bottom) for two conditions are plotted. Two hundred trials are simulated in each condition. (B) The discrete-time hazard functions (top) and conditional accuracy functions (bottom) are plotted for the same data. The first second after target stimulus onset (time zero) is divided in ten bins of 100 ms (t = 1 to 10). The first bin is [0,100], the last bin is [900,1000]. Note that the hazard and conditional accuracy estimates are plotted at the endpoint of each time bin. The definitions of discrete-time hazard and conditional accuracy are further explained in section 2.1.2. Error bars represent +/- 1 standard error of the mean (A) or proportion (B).

@@

To apply discrete-time EHA, one divides the within-trial time in discrete, contiguous time bins indexed by t (e.g., t = 1 to 10; Figure 1B).
Then let RT be a discrete random variable denoting the rank of the time bin in which a particular person’s response occurs in a particular trial across a repeated measures design. 
For example, a response in one trial might occur at 546 ms and it would be in time bin 6 (any RTs from 501 ms to 600 ms).
One then calculates the sample-based estimate of the discrete-time hazard function of event occurrence for each experimental condition (Figure 1B upper panel).
The discrete-time hazard function gives you, for each time bin, the conditional probability that the event occurs (sometime) in
bin t, given that the event does not occur in previous bins. 
In other words, it reflects the instantaneous risk that the event occurs in the current bin t, given that it has not yet
occurred in the past, i.e., in one of the prior bins (t-1, t-2, ..., 1).

@@

@@

In general, it is important to visually inspect the functions first for each participant, in order to identify individuals that may not be following task instructions (e.g., a flat conditional accuracy function at .5 indicates that someone is just guessing), outlying individuals, and/or different groups with qualitatively different behavior. Also, to select a suited bin width for model fitting, one can test and compare various bin widths in the censor function, and select the smallest one that is supported by the data.

@@

The middle column of Supplementary Figure 3 (section F of the Supplemental Material) shows six examples of prior distributions for an intercept on the logit and/or cloglog scales. While a normal distribution with relatively large variance is often used as a weakly informative prior for continuous dependent variables, rows A and B of Supplementary Figure 3 show that specifying such distributions on the logit and cloglog scales actually leads to rather informative distributions on the original probability scale, as most mass is pushed to probabilities of 0 and 1. As such, we modify the prior formulation in order to make sure that it remains consistent with a weakly informative approach (see section F of the Supplemental Material).

@@

Estimating model M1i took about 124 minutes using the same MacBook Pro.

@@

@@

@@

@@

The general approach to planning that we adopt here involves simulating reasonably structured data to help guide what you might be able to expect from your data once you collect it [@gelmanBayesianWorkflow2020].
The basic structure and code follows the examples outlined by Solomon Kurz in his 'power' blog posts (https://solomonkurz.netlify.app/blog/bayesian-power-analysis-part-i/) and Lisa Debruine's R package faux{} (https://debruine.github.io/faux/), as well as these related papers [@debruineUnderstandingMixedEffectsModels2021; @pargentTutorialTailoredSimulationBased2024].

@@

Summary results from Simulation 1 are shown in Figure 6A.
Figure 6A depicts statistical "power" as calculated by the percentage of lower-bound 95% confidence intervals that exclude zero when the difference between prime condition is calculated (congruent - incongruent). 
In other words, we calculate the fraction of simulated datasets that generated an effect of prime that excludes the criterion mark of zero.
We are aware that "power" is not part of a Bayesian analytical workflow, but we choose to include it here, as it is familiar to most researchers in experimental psychology.

@@

But, and this is an important caveat, unless there are unavoidable reasons, no matter what kind of planning choices we made based on these data simulations, we would not solely rely on data collected from one single study. 
Instead, we would run a follow-up experiment that replicates and extends the initial result. 
By doing so, we would aim to avoid the Cult of the Isolated Single Study [@nelderStatisticsStatisticalScience1999; @tongStatisticalInferenceEnables2019], and thus reduce the reliance on any one type of planning tool, such as a power analysis. 
Then, we would look for common patterns across two or more experiments, rather than trying to make the case that a single study on its own has sufficient evidential value to hit some criterion mark.

@@

For those researchers, like ourselves, who are primarily interested in understanding human cognitive and brain systems, we consider two broadly-defined, main use-cases of EHA. First, as we hope to have made clear by this point, EHA is one way to investigating a "temporal states" approach to cognitive processes, by tracking behavior as a function of step-wise increases in absolute waiting time. EHA thus provides a way to uncover the microgenesis of cognitive effects, by revealing when cognitive states may start and stop, how states are replaced with others, as well as what they may be tied to or interact with. Therefore, if your research questions concern **when psychological states occur, and how they are temporally organized**, our EHA tutorials could be useful tools to use for basic knowledge development, as well as theory building.

Second, even if you are not primarily interested in studying the temporal organization of cognitive states, EHA could still be a useful tool to consider using, in order to qualify inferences that are being made based on comparisons between means. Given that distinctly different inferences can be made from the same data based on whether one computes a mean across trials or a RT distribution of events (Figure 1), it may be important for researchers to supplement comparisons between means with EHA. For instance, EHA might reveal that the conclusion of interest based on averaging across trials does not apply to all responses, but is instead restricted to certain periods of within-trial time.
